# Executive Summary

Directed fishing of skates is currently prohibited in the Gulf of Alaska (GOA). However, incidental catches of skates in other fisheries are sufficiently high enough that skates are considered to be “in the fishery” and require harvest specifications. The GOA skate complex is managed as three separate units: big skate (*Beringraja binoculata*), longnose skate (*Raja rhina*) and other skates. Big and Longnose skates have GOA-wide overfishing limits (OFLs) while the Acceptable Biological Catches (ABCs) are region specific (western [WGOA], central [CGOA], and eastern [EGOA]). Remaining skate species are managed as an “other skates” group, with GOA-wide harvest specifications. All GOA skates are managed under Tier 5, where OFL and ABC are based on survey biomass estimates and natural mortality rate (*M*). 

## Summary of Changes in Assessment Inputs

*Changes in the input data:*

1. Updated catch estimates from `r year-1` and `r year` (`r year` catch data as of `r cat_dat_date`)\
2. Biomass estimates from the 2023 GOA bottom trawl survey.

*Changes in the assessment methodology:*\

No changes were made to the assessment methodology.


## Summary of Results

For `r year + 1`, the recommended maximum allowable ABC from the Tier `r tier` specifications for Big skate is `r format(as.numeric(bg_ABC[1,4]), big.mark=",")` t in WGOA, `r format(as.numeric(bg_ABC[2,4]), big.mark=",")` t in CGOA and `r format(as.numeric(bg_ABC[3,4]), big.mark=",")` t in EGOA. The specifications for Longnose skate is `r format(as.numeric(ln_ABC[1,4]), big.mark=",")` t in WGOA, `r format(as.numeric(ln_ABC[2,4]), big.mark=",")` t in CGOA and `r format(as.numeric(ln_ABC[3,4]), big.mark=",")` t in EGOA. The specifications for Other skates is `r format(as.numeric(ot_ABC[4]), big.mark=",")` t. Reference values for all skates are summarized in the following tables, with the recommended ABC and OFL values for `r year + 1` in bold. In the tables below W, C, and E indicate the Western, Central, and Eastern GOA regulatory regions respectively.

```{r, child= "01a-main_table.Rmd", eval = T}

```


None of the three assessment groups experienced overfishing in `r year -1`. This statement was determined by comparing the catch in `r year -1` to the associated OFL. For Big skate, the catch and OFL were `r format(round(big_cur_catch$total_catch[1]), big.mark=",")` t and `r format(round(big_cur_catch$OFL[1]), big.mark=",")` t. For Longnose skate, the catch and OFL were `r format(round(ln_cur_catch$total_catch[1]), big.mark=",")` t and `r format(round(ln_cur_catch$OFL[1]), big.mark=",")` t and for Other skates, the catch and OFL were `r format(round(ot_cur_catch$total_catch[1]), big.mark=",")` t and `r format(round(ot_cur_catch$OFL[1]), big.mark=",")` t. 

As of `r cat_dat_date`, none of the assessments groups are experiencing overfishing; Big skate catch and OFL are `r format(round(big_cur_catch$total_catch[2]), big.mark=",")` t and `r format(round(big_cur_catch$OFL[2]), big.mark=",")` t, Longnose skate catch and OFL are `r format(round(ln_cur_catch$total_catch[2]), big.mark=",")` t and `r format(round(ln_cur_catch$OFL[2]), big.mark=",")` t and Other skate catch and OFL are `r format(round(ot_cur_catch$total_catch[2]), big.mark=",")` t and `r format(round(ot_cur_catch$OFL[2]), big.mark=",")` t.  As for `r year` ABCs, Big and Other skates have catches below their regional ABCs (Tables `r params$chapter`-\@ref(tab:Tab-big-cat-ofl) and -\@ref(tab:Tab-other-cat-ofl)). Longnose skate has catches below their regional ABCs except for EGOA where the catch is larger than teh regional ABC (`r format(round(ln_cur_catch$EGOA[2]), big.mark=",")` t catch and `r format(round(ln_cur_catch$E_ABC[2]), big.mark=",")` t ABC) (Table `r params$chapter`-\@ref(tab:Tab-longnose-cat-ofl)). 


```{r, child= "01b-apportionment.Rmd", eval = F}

```


```{r, child= "01c-plan_team.Rmd", eval = F}

```


